---
title: "`QuIBL` introgression analysis of _Solenopsis_ species (17 samples)"
author: "Federico López-Osorio"
date: '`r Sys.Date()`'
output:
  pdf_document: 
    fig_caption: yes
    toc: yes
  html_document:
    toc: yes
editor_options: 
  chunk_output_type: console
geometry: margin = 1cm
---

```{r setup, echo = FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  cache.lazy = TRUE,
  include = TRUE
  # out.height = "\textheight",
  # out.width = "\textwidth"
)
```

# Load libraries
```{r}
load_cran_pkgs <- function(pkg) {
  new_pkg <- pkg[!(pkg %in% installed.packages()[, "Package"])]
  if (length(new_pkg)) {
    install.packages(new_pkg, dependencies = TRUE)
  }
  sapply(pkg, require, character.only = TRUE)
}

# Install packages available on CRAN
cran_pkgs <- c(
  "devtools", "tidyverse", "RColorBrewer", "here", "styler",
  "ggrepel", "magrittr", "ape", "hash", "ggpubr", "egg", "ggpubr"
)
load_cran_pkgs(cran_pkgs)

# Install packages available on Bioconductor
load_bioconductor_pkgs <- function(pkg) {
  new_pkg <- pkg[!(pkg %in% installed.packages()[, "Package"])]
  if (length(new_pkg)) {
    BiocManager::install(new_pkg, version = "3.13")
  }
  sapply(pkg, require, character.only = TRUE)
}

bioconductor_pkgs <- c("ggtree")

load_bioconductor_pkgs(bioconductor_pkgs)

# Install "quiblR"
devtools::install_github("nbedelman/quiblR")
library(quiblR)
```

# Set a custom `ggplot` theme
```{r}
# Set a custom ggplot theme
# devtools::install_github("cttobin/ggthemr")
library(ggthemr)
custom_palette <- c(
  "#9E9E9E", "#59BDEF", "#EBCC2A", "#E1AF00", "#F27C8D", "#7B73D1", "#7B9FE0",
  "#F5CC7F", "#66C2A5", "#28A7B6", "#F2CB57", "#F2A057", "#F2845C"
)
# ggthemr::colour_plot(custom_palette)

custom_theme <- ggthemr::define_palette(
  swatch = custom_palette,
  gradient = c(lower = "#59BDEF", upper = "#FF6B5A")
)
ggthemr::ggthemr(custom_theme)
```

The analysis below follows the procedures documented in the `quiblR` package
(https://github.com/nbedelman/quiblR).

# Load data
`quiblR` requires three input files:  

* A species tree  
* QuIBL output  
* A list of gene trees  

```{r}
species_tree <- quiblR::read_speciesTree(
  here::here("data", "Astral.10SNP-genes.chr1-15.nwk")
)

quibl_output <- quiblR::read_csv_quibl(
  here::here("data", "chr16nr.17samples.quibl.out.csv")
)

original_trees <- ape::read.tree(
  here::here("data", "chr16nr.17samples.raxml.rooted.bestTree")
)

length(original_trees)
```

# Replace underscores with dashes in sample names
```{r}
# get tip names from one of the subset trees
subset_tips <- c(original_trees[[1]]$tip.label)

# subset the large species tree to include only the tips of interest
species_tree <- ape::keep.tip(species_tree, subset_tips)

# replace underscores with dashes
# QuIBL and quiblR use underscores to separate triplets
species_tree$tip.label <- gsub(
  "(SRR.+?)_", "\\1-", species_tree$tip.label
)

quibl_output <- quibl_output %>%
  dplyr::mutate_at(c("triplet", "outgroup"),
    stringr::str_replace_all,
    pattern = "(SRR.+?)_", replacement = "\\1-"
  )

original_trees <- lapply(original_trees, function(x) {
  x$tip.label <- gsub("(SRR.+?)_", "\\1-", x$tip.label)
  return(x)
})

class(original_trees) <- "multiPhylo"

# plot tree including the subset of tips
# ape::plot.phylo(species_tree)
pruned_tree <- ggtree(species_tree,
  size = 1, color = "#767676", ladderize = FALSE
) +
  ggtree::geom_tiplab(size = 4, offset = .1, color = "#767676") +
  xlim(0, 28)

pruned_tree

ggsave(pruned_tree,
  filename = "Astral.10SNP-genes.chr1-15.pruned17samples.pdf",
  width = 5,
  height = 8,
  units = "in",
  dpi = "retina"
)
```

# Get big-picture results
QuIBL assumes that we have a rooted species tree. In this case, we have rooted 
the tree by fixing "gem-1-bigB-m-majorityallele" as the outgroup. 

QuIBL discards all triplets that include the overall outgroup, since all loci 
are forced to have the same topology ("gem-1-bigB-m-majorityallele" as the 
outgroup).

We only accept the two-distribution model if "BIC2Dist" is at least 10 units 
lower than "BIC1Dist" (`quiblR::testSignificance()`).

```{r}
ape::is.rooted(species_tree)

# species_tree <- root(species_tree,
#   outgroup = "gem-1-bigB-m-majorityallele", resolve.root = TRUE
# )

# check tip labels
species_tree$tip.label %>% head()
original_trees[[1]]$tip.label %>% head()

# sanity check: compare tip labels
setdiff(species_tree$tip.label, original_trees[[1]]$tip.label)

# identify topologies concordant with the species tree and topologies which
# represent either ILS or introgression
# identify significant differences between the two models
total_trees <- sum(quibl_output$count) / length(unique(quibl_output$triplet))

quibl_output <- dplyr::mutate(quibl_output,
  isDiscordant = as.integer(!apply(quibl_output, 1,
    quiblR::isSpeciesTree,
    sTree = species_tree
  )),
  isSignificant = as.integer(apply(quibl_output, 1,
    quiblR::testSignificance,
    threshold = 10
  )),
  totalIntrogressionFraction = (mixprop2 * count * isDiscordant) / total_trees
)

head(quibl_output)

# save the results including discordant and significant triplets
quibl_output_sig <- quibl_output %>%
  dplyr::filter(isDiscordant == 1 & isSignificant == 1) %>%
  dplyr::arrange(desc(totalIntrogressionFraction))

head(quibl_output_sig)
# 6 significant results using the default "threshold = 10"
# high total introgression fraction (63.6%) between
# "GCa8-9-littleb-p_GCa3-9-littleb-p"
# (with outgroup "SRR9008214-meg-130-BZ-bigB")

write.csv(quibl_output_sig,
  file = here::here("chr16nr_17samples_quiblr_introgression_sig.csv"),
  row.names = FALSE
)
```

# Get summary of results
`quiblR`'s `getIntrogressionSummary()` function returns a data frame with the 
average introgression fraction for each pair of taxa. Specifically, the function
goes triplet by triplet, ignores topologies concordant with the species tree, 
records `mixprop2 * count / total trees`, and averages that value across all
tests that include each pair of species.

```{r}
introgression_summary <- quiblR::getIntrogressionSummary(
  quibl_output, species_tree
)

introgression_summary <- introgression_summary %>%
  dplyr::arrange(desc(value))

# save the introgression summary
write.csv(introgression_summary,
  file = here::here("chr16nr_17samples_quiblr_introgression_summary.csv"),
  row.names = FALSE
)

head(introgression_summary)
```

# Make a heatmap of the average introgression fractions
```{r}
# plot heatmap of average introgression fractions
str(introgression_summary)

# create a function to plot heatmaps of average introgression fractions
plot_heatmap <- function(df) {
  ggplot(data = df, aes(x = tax1, y = tax2, fill = value)) +
    geom_tile(color = "white") +
    scale_fill_gradient2(
      low = "#3B9AB2", high = "#F21A00", mid = "#EBCC2A", na.value = "grey50",
      midpoint = max(introgression_summary$value) / 2,
      limit = c(0, max(introgression_summary$value)),
      name = "Average introgression fraction"
    ) +
    geom_abline(slope = 1, intercept = 0, color = "grey50") +
    geom_vline(
      xintercept = seq(1.5, nrow(introgression_summary) + 0.5, 1),
      alpha = 0.6
    ) +
    geom_hline(
      yintercept = seq(1.5, nrow(introgression_summary) + 0.5, 1),
      alpha = 0.6
    ) +
    labs(x = "", y = "") +
    # scale_x_discrete(position = "top") +
    theme(
      panel.grid = element_blank(),
      # legend.position = "none",
      legend.position = "top",
      legend.title.align = 0.5,
      legend.box = "vertical",
      legend.margin = margin(),
      text = element_text(color = "#767676"),
      # legend.title = element_text(size = 12),
      # legend.text = element_text(size = 10),
      axis.text = element_text(color = "#767676"),
      axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1),
      plot.margin = margin(6, 6, 0, 0)
    )
}

summary_matrix <- plot_heatmap(introgression_summary)
summary_matrix

species_tree_subset <- ggtree(quiblR::extractTripletTree(
  species_tree,
  unique(introgression_summary$tax1)
),
size = 1, color = "#767676", ladderize = FALSE,
layout = "rect", branch.length = "none"
) +
  theme(plot.margin = margin(0, 0, 0, 0))
# ggtree::geom_tiplab(size = 4, offset = .5, color = "#767676") +
# xlim(0, 32)
# species_tree_subset

species_tree_subset_down <- ggtree(quiblR::extractTripletTree(
  species_tree,
  unique(introgression_summary$tax1)
),
size = 1, color = "#767676", ladderize = FALSE,
layout = "rect", branch.length = "none"
) +
  coord_flip() +
  theme(plot.margin = margin(0, 0, 0, 0))
# ggtree::geom_tiplab(
#   angle = 90, vjust = 0.5, hjust = 0, size = 4, offset = .5,
#   color = "#767676"
# ) +
# xlim(0, 46)
# species_tree_subset_down
# geom_tiplab(angle = 90, vjust = 0.5, hjust = 1)

empty <- ggplot() +
  theme_void()

introgression_heatmap <- egg::ggarrange(
  species_tree_subset, summary_matrix, empty, species_tree_subset_down,
  ncol = 2, nrow = 2, heights = c(2, 1), widths = c(1, 2)
)

ggsave(introgression_heatmap,
  filename = "chr16nr_17samples_introgression_heatmap.pdf",
  width = 8,
  height = 8,
  units = "in",
  dpi = "retina"
)

# introgression_heatmap <- ggpubr::ggarrange(
#   speciesTreeSubset, summaryMatrix, NULL, speciesTreeSubset_down,
#   ncol = 2, nrow = 2, align = "hv", heights = c(2, 1), widths = c(1, 2),
#   common.legend = TRUE
# )

# plot heatmap with samples sorted according to species and supergene haplotype
unique_tax <- introgression_summary %>%
  dplyr::pull(tax1) %>%
  unique() %>%
  as.character()

ordered_samples <- as.data.frame(unique_tax) %>%
  dplyr::rename(sample = unique_tax) %>%
  dplyr::mutate(
    species = case_when(
      grepl("-inv-", sample) ~ "invicta",
      grepl("-ric-", sample) ~ "richteri",
      grepl("-xAdR-", sample) ~ "xAdR",
      grepl("-meg-|GCa", sample) ~ "megergates",
      grepl("-int-", sample) ~ "interrupta",
      grepl("-sae-", sample) ~ "saevissima",
      grepl("Lad4", sample) ~ "pusillignis"
    ),
    genotype = case_when(
      grepl("bigB", sample) ~ "bigB",
      grepl("littleb", sample) ~ "littleb"
    )
  ) %>%
  dplyr::arrange(species, genotype)

introgression_summary_ordered <- introgression_summary

# reorder samples according to species and supergene form
introgression_summary_ordered$tax1 <- factor(
  introgression_summary_ordered$tax1,
  levels = rev(ordered_samples$sample)
  # levels = rev(ordered_samples$sample)
)

introgression_summary_ordered$tax2 <- factor(
  introgression_summary_ordered$tax2,
  levels = rev(ordered_samples$sample)
)

plot_heatmap(introgression_summary_ordered) +
  geom_rug(
    data = ordered_samples,
    aes(x = sample, color = species),
    length = unit(0.01, "npc"), size = 4, sides = "b",
    inherit.aes = FALSE
  ) +
  geom_rug(
    data = ordered_samples,
    aes(y = sample, color = species),
    length = unit(0.01, "npc"), size = 4, sides = "l",
    inherit.aes = FALSE
  ) +
  scale_x_discrete(expand = expansion(mult = c(0.06, 0))) +
  scale_y_discrete(expand = expansion(mult = c(0.06, 0))) +
  guides(color = guide_legend(title = "Species")) +
  ggthemr::scale_colour_ggthemr_d()

ggsave(
  filename = "chr16nr_17samples_introgression_heatmap_ordered.pdf",
  width = 10,
  height = 10,
  units = "in",
  dpi = "retina"
)
```

# Examine triplets of interest
```{r}
# select a triplet of interest
head(quibl_output_sig)

quibl_output %>%
  dplyr::filter(isDiscordant == 1 & isSignificant == 1) %>%
  dplyr::arrange(desc(totalIntrogressionFraction)) %>%
  head(n = 20)

target_triplet <-
  "SRR9008214-meg-130-BZ-bigB_GCa8-9-littleb-p_GCa3-9-littleb-p"

quibl_output %>%
  dplyr::filter(stringr::str_detect(triplet, target_triplet))

# subset(quibl_output, triplet == target_triplet)
list_target_triplets <- list(
  "SRR9008214-meg-130-BZ-bigB_GCa8-9-littleb-p_GCa3-9-littleb-p",
  "SRR9008232-xAdR-134-Bra-bigB_GCa8-9-littleb-p_GCa3-9-littleb-p",
  "GCa3-11-bigB-p_GCa8-9-littleb-p_GCa3-9-littleb-p",
  "SRR9008232-xAdR-134-Bra-bigB_SRR9008163-inv-225-Mis-littleb_GCa3-9-littleb-p"
)
```

`getPerLocusStats()` produces a data frame including the triplet sub-tree, 
the outgroup, internal branch length, and introgression probability.
```{r}
triplet_perlocus_stats <- quiblR::getPerLocusStats(
  quiblOutput = quibl_output,
  trip = target_triplet,
  treeList = original_trees,
  overallOut = "gem-1-bigB-m-majorityallele"
)

head(triplet_perlocus_stats)

# plot the introgression probabilities across loci
triplet_perlocus_stats %>%
  ggplot(aes(x = introProb)) +
  geom_histogram(fill = "grey50", bins = 50)

multiple_triplets_perlocus_stats <- list()

for (i in list_target_triplets) {
  multiple_triplets_perlocus_stats[[i]] <- quiblR::getPerLocusStats(
    quiblOutput = quibl_output,
    trip = i,
    treeList = original_trees,
    overallOut = "gem-1-bigB-m-majorityallele"
  )
}
```

Look at the internal branch lengths for introgression topologies.
```{r}
plot_histo_blens <- function(df, outgroup) {
  ggplot(
    data = subset(
      df, out == outgroup
    ),
    aes(x = branchLength)
  ) +
    geom_histogram(fill = "grey50", bins = 80) +
    labs(title = target_triplet, x = "Branch Length", y = "Count") +
    theme(
      plot.title = element_text(hjust = 0.5, size = 14, face = "plain"),
      axis.title = element_text(size = 14),
      axis.text = element_text(size = 12)
    ) +
    xlim(0, 0.005)
}

plot_density_blens <- function(df, outgroup, target_triplet) {
  ggplot(
    data = subset(
      df, out == outgroup
    ),
    aes(x = branchLength)
  ) +
    geom_density(fill = "grey50", color = "grey50", alpha = 0.6) +
    labs(title = target_triplet, x = "Branch Length", y = "Density") +
    theme(
      plot.title = element_text(hjust = 0.5, size = 14, face = "plain"),
      axis.title = element_text(size = 14),
      axis.text = element_text(size = 12)
    ) +
    xlim(0, 0.005)
}

names(multiple_triplets_perlocus_stats)
# [1] "SRR9008214-meg-130-BZ-bigB_GCa8-9-littleb-p_GCa3-9-littleb-p"
# [2] "SRR9008232-xAdR-134-Bra-bigB_GCa8-9-littleb-p_GCa3-9-littleb-p"
# [3] "GCa3-11-bigB-p_GCa8-9-littleb-p_GCa3-9-littleb-p"
# [4] "SRR9008232-xAdR-134-Bra-bigB_SRR9008163-inv-225-Mis-littleb_GCa3-9-littleb-p"

pd1 <- plot_density_blens(
  multiple_triplets_perlocus_stats[[1]],
  "SRR9008214-meg-130-BZ-bigB",
  names(multiple_triplets_perlocus_stats)[1]
)

pd2 <- plot_density_blens(
  multiple_triplets_perlocus_stats[[2]],
  "SRR9008232-xAdR-134-Bra-bigB",
  names(multiple_triplets_perlocus_stats)[2]
)

pd3 <- plot_density_blens(
  multiple_triplets_perlocus_stats[[3]],
  "GCa3-11-bigB-p",
  names(multiple_triplets_perlocus_stats)[3]
)

pd4 <- plot_density_blens(
  multiple_triplets_perlocus_stats[[4]],
  "SRR9008232-xAdR-134-Bra-bigB",
  names(multiple_triplets_perlocus_stats)[4]
)

density_triplets <- gridExtra::grid.arrange(pd1, pd2, pd3, pd4, ncol = 2)

ggsave(
  plot = density_triplets,
  filename = "chr16nr_17samples_select_triplets_blengths_density.pdf",
  width = 8,
  height = 8,
  units = "in",
  dpi = "retina"
)
```
